---
title: "General IPM with Seed Bank"
author: "Simon Frost"
format:
  html:
    code-fold: false
    toc: true
---

## Overview

A **general IPM** models populations with multiple state variables — both continuous (e.g., size) and discrete (e.g., seed bank, dormancy stage). The kernel becomes a block-structured **mega-kernel** with transitions between all state combinations.

This vignette models *Ligustrum obtusifolium* (border privet) with a discrete seed bank and continuous height, following Levin et al. (2019). We compare two treatments: **control** and **crown removal (CR)**.

## Setup

```{r}
library(ipmr)
```

## The Ligustrum Model

The population has two states:

- **Height** (`ht`): continuous, measured in cm — domain $[1.02, 624]$
- **Seed bank** (`b`): discrete, a single pool of seeds in the soil

The mega-kernel has four blocks:

$$
K = \begin{pmatrix}
K_{CC} & K_{DC} \\
K_{CD} & K_{DD}
\end{pmatrix}
$$

where:

- $K_{CC}$ (continuous → continuous): survival-growth (P kernel)
- $K_{CD}$ (continuous → discrete): seeds entering the seed bank
- $K_{DC}$ (discrete → continuous): seedlings emerging from the seed bank
- $K_{DD}$ (discrete → discrete): seeds staying in the seed bank (= 0 in this model)

## Parameters

### Control Treatment

```{r}
ctrl <- list(
  g_int     = 5.781,
  g_slope   = 0.988,
  g_sd      = 20.55699,
  s_int     = -0.352,
  s_slope   = 0.122,
  s_slope_2 = -0.000213,
  f_r_int   = -11.46,
  f_r_slope = 0.0835,
  f_s_int   = 2.6204,
  f_s_slope = 0.01256,
  f_d_mu    = 5.6655,
  f_d_sd    = 2.0734,
  e_p       = 0.15,
  g_i       = 0.5067
)
```

### CR Treatment

```{r}
cr <- list(
  g_int     = 7.229,
  g_slope   = 0.988,
  g_sd      = 21.72262,
  s_int     = 0.0209,
  s_slope   = 0.0831,
  s_slope_2 = -0.00012999,
  f_r_int   = -11.46,
  f_r_slope = 0.0835,
  f_s_int   = 2.6204,
  f_s_slope = 0.01256,
  f_d_mu    = 5.6655,
  f_d_sd    = 2.0734,
  e_p       = 0.15,
  g_i       = 0.5067
)
```

## Domain

```{r}
L <- 1.02
U <- 624.0
n <- 500
```

## Building the General IPM

In `ipmr`, a general model uses `sim_gen = "general"`. The `family` argument specifies the transition type: `"CC"` (continuous→continuous), `"CD"` (continuous→discrete), `"DC"` (discrete→continuous), or `"DD"` (discrete→discrete). Note that in general models, the mesh width `d_ht` must be explicitly included in the formula.

We define a helper function to build both treatments with the same structure.

```{r}
inv_logit_2 <- function(int, slope, slope_2, z) {
  1 / (1 + exp(-(int + slope * z + slope_2 * z^2)))
}

build_ligustrum <- function(params) {
  states <- list(c("ht", "b"))

  init_ipm(sim_gen = "general", di_dd = "di", det_stoch = "det") %>%
    # CC: survival-growth (height → height)
    define_kernel(
      name      = "P",
      formula   = s * g * d_ht,
      family    = "CC",
      s         = inv_logit_2(s_int, s_slope, s_slope_2, ht_1),
      g         = dnorm(ht_2, g_mu, g_sd),
      g_mu      = g_int + g_slope * ht_1,
      data_list = params,
      states    = states,
      evict_cor = TRUE,
      evict_fun = truncated_distributions("norm", "g")
    ) %>%
    # CD: continuous → discrete (seeds entering bank)
    define_kernel(
      name      = "go_discrete",
      formula   = f_r * f_s * g_i,
      family    = "CD",
      f_r       = 1 / (1 + exp(-(f_r_int + f_r_slope * ht_1))),
      f_s       = exp(f_s_int + f_s_slope * ht_1),
      data_list = params,
      states    = states,
      evict_cor = FALSE
    ) %>%
    # DD: discrete → discrete (stasis in bank, = 0)
    define_kernel(
      name      = "stay_discrete",
      formula   = 0,
      family    = "DD",
      states    = states,
      evict_cor = FALSE
    ) %>%
    # DC: discrete → continuous (seedlings emerging)
    define_kernel(
      name      = "leave_discrete",
      formula   = e_p * f_d * d_ht,
      family    = "DC",
      f_d       = dnorm(ht_2, f_d_mu, f_d_sd),
      data_list = params,
      states    = states,
      evict_cor = TRUE,
      evict_fun = truncated_distributions("norm", "f_d")
    ) %>%
    define_impl(
      make_impl_args_list(
        kernel_names = c("P", "go_discrete", "stay_discrete", "leave_discrete"),
        int_rule     = rep("midpoint", 4),
        state_start  = c("ht", "ht", "b",  "b"),
        state_end    = c("ht", "b",  "b",  "ht")
      )
    ) %>%
    define_domains(ht = c(L, U, n)) %>%
    define_pop_state(
      pop_vectors = list(
        n_ht = rep(1 / n, n),
        n_b  = 20
      )
    ) %>%
    make_ipm(iterations = 100, usr_funs = list(inv_logit_2 = inv_logit_2))
}
```

### Build Both Models

```{r}
ipm_ctrl <- build_ligustrum(ctrl)
ipm_cr   <- build_ligustrum(cr)
```

## Eigenanalysis

```{r}
lam_ctrl <- lambda(ipm_ctrl)
lam_cr   <- lambda(ipm_cr)
```

```{r}
#| echo: false
cat("Control treatment:\n")
cat("  λ =", round(lam_ctrl, 4), "\n")
cat("\nCrown removal treatment:\n")
cat("  λ =", round(lam_cr, 4), "\n")
```

### Published Values

The published values from Levin et al. (2019) are:

- Control: $\lambda \approx 1.22$
- CR: $\lambda \approx 1.26$

```{r}
#| echo: false
cat("Comparison with published values:\n")
cat(sprintf("  Control: computed=%.2f, published=1.22, diff=%.3f\n",
            lam_ctrl, abs(lam_ctrl - 1.22)))
cat(sprintf("  CR:      computed=%.2f, published=1.26, diff=%.3f\n",
            lam_cr, abs(lam_cr - 1.26)))
```

## Visualizing the Mega-Kernel

```{r}
h_val <- (U - L) / n
meshpts <- L + ((1:n) - 1/2) * h_val

# Extract CC block for visualization
P_ctrl <- ipm_ctrl$sub_kernels$P

image(meshpts, meshpts, t(P_ctrl),
      xlab = "Height (t)", ylab = "Height (t+1)",
      main = "CC block: survival-growth (control)",
      col = hcl.colors(50, "viridis"))
```

## Stable Distribution

```{r}
w_ctrl <- right_ev(ipm_ctrl)

# Height distribution
w_ht <- w_ctrl$ht_w / sum(c(w_ctrl$ht_w, w_ctrl$b_w))
w_b  <- sum(w_ctrl$b_w) / sum(c(w_ctrl$ht_w, w_ctrl$b_w))
```

```{r}
#| echo: false
cat("Proportion in height classes:", round(sum(w_ht), 4), "\n")
cat("Proportion in seed bank:", round(w_b, 4), "\n")
```

```{r}
plot(meshpts, w_ht, type = "l", lwd = 2, col = "blue",
     xlab = "Height (cm)", ylab = "Proportion",
     main = "Stable height distribution (control)")
polygon(c(meshpts, rev(meshpts)),
        c(w_ht, rep(0, n)),
        col = adjustcolor("blue", alpha.f = 0.3), border = NA)
lines(meshpts, w_ht, lwd = 2, col = "blue")
```

## Vital Rate Diagnostics

```{r}
surv_ctrl <- function(z) 1 / (1 + exp(-(ctrl$s_int + ctrl$s_slope * z + ctrl$s_slope_2 * z^2)))
surv_cr   <- function(z) 1 / (1 + exp(-(cr$s_int + cr$s_slope * z + cr$s_slope_2 * z^2)))
repr_prob <- function(z) 1 / (1 + exp(-(ctrl$f_r_int + ctrl$f_r_slope * z)))
seed_prod <- function(z) exp(ctrl$f_s_int + ctrl$f_s_slope * z)

z_seq <- seq(L, U, length.out = 200)
z_off <- seq(0, 30, length.out = 200)

par(mfrow = c(2, 2))
plot(z_seq, sapply(z_seq, surv_ctrl), type = "l", lwd = 2,
     xlab = "Height", ylab = "Probability", main = "Quadratic survival")
lines(z_seq, sapply(z_seq, surv_cr), lwd = 2, lty = 2)
legend("topright", c("Control", "CR"), lty = c(1, 2), lwd = 2, cex = 0.8)

plot(z_seq, sapply(z_seq, repr_prob), type = "l", lwd = 2,
     xlab = "Height", ylab = "Probability", main = "Reproduction probability")

plot(z_seq, sapply(z_seq, seed_prod), type = "l", lwd = 2,
     xlab = "Height", ylab = "Seeds", main = "Seed production")

plot(z_off, dnorm(z_off, ctrl$f_d_mu, ctrl$f_d_sd), type = "l", lwd = 2,
     xlab = "Offspring height", ylab = "Density", main = "Offspring size distribution")
par(mfrow = c(1, 1))
```

## Population Dynamics

```{r}
all_lambdas_ctrl <- lambda(ipm_ctrl, type_lambda = "all")

plot(all_lambdas_ctrl, type = "l", lwd = 2,
     xlab = "Time step", ylab = expression(lambda(t)),
     main = "Per-step growth rate (control)")
abline(h = lam_ctrl, col = "red", lty = 2, lwd = 2)
legend("topright", c(expression(lambda(t)), expression("Asymptotic " * lambda)),
       lty = c(1, 2), col = c("black", "red"), lwd = 2)
```

## Comparing Treatments

```{r}
barplot(c(lam_ctrl, lam_cr), names.arg = c("Control", "Crown Removal"),
        ylab = expression(lambda),
        main = "Treatment comparison",
        col = c("steelblue", "coral"),
        ylim = c(1.0, 1.35))
abline(h = 1.0, lty = 2, lwd = 2)
```

The crown removal treatment has a *higher* $\lambda$ than the control, suggesting that the removal of competing crown cover facilitates Ligustrum invasion.

## Block Structure Explained

The mega-kernel structure:

```
            Current state
            ht (500)    b (1)
          ┌──────────┬────────┐
Next  ht  │  CC: P   │  DC:   │
state     │  s×g     │  e_p×  │
(500)     │  (500×500)│  f_d   │
          │          │  (500×1)│
          ├──────────┼────────┤
      b   │  CD:     │  DD:   │
     (1)  │  f_r×f_s │   0    │
          │  ×g_i    │  (1×1) │
          │  (1×500) │        │
          └──────────┴────────┘
```

- **CC** ($500 \times 500$): Surviving plants grow (normal growth with truncation)
- **CD** ($1 \times 500$): Reproducing plants send seeds to the bank ($f_r \times f_s \times g_i$)
- **DC** ($500 \times 1$): Seeds germinate and establish as seedlings ($e_p \times f_d$)
- **DD** ($1 \times 1$): Seeds do not stay in the bank (= 0 in this model)

## Summary

In this vignette we:

1. Built a **general IPM** with a continuous height state and a discrete seed bank
2. Used `ipmr`'s general model with CC, CD, DC, and DD kernel families
3. Applied `truncated_distributions` eviction correction to growth and offspring size
4. Computed $\lambda$ for control ($\approx 1.22$) and crown removal ($\approx 1.26$) treatments
5. Validated against published results from Levin et al. (2019)
