---
title: "Sensitivity and Elasticity Analysis"
author: "Simon Frost"
format:
  html:
    code-fold: false
    toc: true
---

## Overview

**Perturbation analysis** quantifies how the population growth rate $\lambda$ responds to changes in the projection kernel. Two key measures are:

- **Sensitivity**: $\frac{\partial \lambda}{\partial K(z', z)} = \frac{v(z') \, w(z)}{\langle v, w \rangle}$ — the absolute change in $\lambda$ for a small additive change to $K(z', z)$
- **Elasticity**: $e(z', z) = \frac{K(z', z)}{\lambda} \cdot \frac{\partial \lambda}{\partial K(z', z)}$ — the proportional change in $\lambda$ for a proportional change to $K(z', z)$

A fundamental property is that the **elasticities sum to 1**, and the elasticities of the P and F kernels sum to the respective contributions to $\lambda$.

This vignette uses the Soay sheep (ungulate) model from the previous vignette.

## Setup

```{r}
library(ipmr)
```

## Soay Sheep Model

```{r}
data_list <- list(
  surv_int  = -9.65,
  surv_z    =  3.77,
  grow_int  =  1.41,
  grow_z    =  0.557,
  grow_sd   =  0.0799,
  repr_int  = -7.23,
  repr_z    =  2.60,
  recr_int  =  1.93,
  recr_pr   =  1 / (1 + exp(-1.93)),  # pre-computed recruitment probability
  rcsz_int  =  0.362,
  rcsz_z    =  0.709,
  rcsz_sd   =  0.159
)

L <- 1.6
U <- 3.7
m <- 100
```

```{r}
ungulate_ipm <- init_ipm(sim_gen = "simple", di_dd = "di", det_stoch = "det") %>%
  define_kernel(
    name          = "P",
    formula       = s * g,
    family        = "CC",
    s             = 1 / (1 + exp(-(surv_int + surv_z * z_1))),
    g             = dnorm(z_2, grow_mu, grow_sd),
    grow_mu       = grow_int + grow_z * z_1,
    data_list     = data_list,
    states        = list(c("z")),
    uses_par_sets = FALSE,
    evict_cor     = FALSE
  ) %>%
  define_kernel(
    name          = "F",
    formula       = s * pb * 0.5 * recr_pr * c_z1z,
    family        = "CC",
    s             = 1 / (1 + exp(-(surv_int + surv_z * z_1))),
    pb            = 1 / (1 + exp(-(repr_int + repr_z * z_1))),
    c_z1z         = dnorm(z_2, rcsz_mu, rcsz_sd),
    rcsz_mu       = rcsz_int + rcsz_z * z_1,
    data_list     = data_list,
    states        = list(c("z")),
    uses_par_sets = FALSE,
    evict_cor     = FALSE
  ) %>%
  define_impl(
    make_impl_args_list(
      kernel_names = c("P", "F"),
      int_rule     = rep("midpoint", 2),
      state_start  = rep("z", 2),
      state_end    = rep("z", 2)
    )
  ) %>%
  define_domains(z = c(L, U, m)) %>%
  define_pop_state(n_z = rep(1 / m, m)) %>%
  make_ipm(iterations = 100)

lam <- lambda(ungulate_ipm)
```

```{r}
#| echo: false
cat("Lambda =", lam, "\n")
```

## Sensitivity Matrix

The sensitivity matrix $S(z', z)$ gives the absolute effect of a small additive change to $K(z', z)$ on $\lambda$:

$$
S(z', z) = \frac{v(z') \, w(z)}{\langle v, w \rangle}
$$

where $w$ is the stable distribution (right eigenvector) and $v$ is the reproductive value (left eigenvector).

```{r}
h <- (U - L) / m
meshpts <- L + ((1:m) - 1/2) * h

# Extract eigenvectors
w_raw <- right_ev(ungulate_ipm)$z_w
v_raw <- left_ev(ungulate_ipm)$z_v

# Normalize: w sums to 1, v scaled so <v, w> = 1
w <- w_raw / sum(w_raw)
v <- v_raw

# Sensitivity matrix: S = v %*% t(w) / <v, w>
vw_inner <- sum(v * w)
S <- outer(v, w) / vw_inner

image(meshpts, meshpts, t(S),
      xlab = "Size at time t", ylab = "Size at time t+1",
      main = "Sensitivity matrix S(z', z)",
      col = hcl.colors(50, "viridis"))
```

```{r}
par(mfrow = c(1, 2))
plot(meshpts, w, type = "l", lwd = 2, col = "blue",
     xlab = "Size", ylab = "w(z)", main = "Stable distribution")
plot(meshpts, v, type = "l", lwd = 2, col = "red",
     xlab = "Size", ylab = "v(z)", main = "Reproductive value")
par(mfrow = c(1, 1))
```

## Elasticity Matrix

The elasticity matrix gives the *proportional* sensitivity — the proportional change in $\lambda$ for a proportional change in $K(z', z)$:

$$
e(z', z) = \frac{K(z', z)}{\lambda} \cdot S(z', z)
$$

```{r}
K_matrix <- ungulate_ipm$sub_kernels$P + ungulate_ipm$sub_kernels$F
E <- (K_matrix / lam) * S

image(meshpts, meshpts, t(E),
      xlab = "Size at time t", ylab = "Size at time t+1",
      main = "Elasticity matrix e(z', z)",
      col = hcl.colors(50, "viridis"))
```

### Elasticities Sum to 1

A key property of the elasticity matrix is that its elements sum to 1 (when scaled by $h^2$ for the double integral over the continuous domain):

In the discretized version, the sum of all matrix elements equals 1:

```{r}
#| echo: false
cat("Sum of elasticities:", sum(E), "\n")
```

## Decomposing P and F Contributions

We can compute the elasticity contributions from the P (survival-growth) and F (fecundity) kernels separately.

```{r}
P_matrix <- ungulate_ipm$sub_kernels$P
F_matrix <- ungulate_ipm$sub_kernels$F

E_P <- (P_matrix / lam) * S
E_F <- (F_matrix / lam) * S

e_P <- sum(E_P)
e_F <- sum(E_F)
```

```{r}
#| echo: false
cat("Elasticity contribution of P (survival-growth):", round(e_P, 6), "\n")
cat("Elasticity contribution of F (fecundity):      ", round(e_F, 6), "\n")
cat("Sum:                                           ", round(e_P + e_F, 6), "\n")
```

```{r}
par(mfrow = c(1, 2))
image(meshpts, meshpts, t(E_P),
      xlab = "Size (t)", ylab = "Size (t+1)",
      main = "Elasticity of P",
      col = hcl.colors(50, "viridis"))
image(meshpts, meshpts, t(E_F),
      xlab = "Size (t)", ylab = "Size (t+1)",
      main = "Elasticity of F",
      col = hcl.colors(50, "viridis"))
par(mfrow = c(1, 1))
```

## Biological Interpretation

For Soay sheep, most of the elasticity is in the P kernel (survival-growth), meaning $\lambda$ is more sensitive to proportional changes in survival and growth than to proportional changes in fecundity. This is typical for long-lived iteroparous species.

### Size-Specific Elasticity

We can marginalize the elasticity matrix to see which sizes contribute most:

```{r}
# Marginal elasticity over z' (contribution of size z at time t)
e_by_size <- colSums(E)

# Marginal elasticity over z (contribution of size z' at time t+1)
e_by_size_prime <- rowSums(E)

par(mfrow = c(1, 2))
plot(meshpts, e_by_size, type = "l", lwd = 2, col = "blue",
     xlab = "Size z (time t)", ylab = "Marginal elasticity",
     main = "Elasticity by current size")
polygon(c(meshpts, rev(meshpts)),
        c(e_by_size, rep(0, m)),
        col = adjustcolor("blue", alpha.f = 0.3), border = NA)
lines(meshpts, e_by_size, lwd = 2, col = "blue")

plot(meshpts, e_by_size_prime, type = "l", lwd = 2, col = "orange",
     xlab = "Size z' (time t+1)", ylab = "Marginal elasticity",
     main = "Elasticity by future size")
polygon(c(meshpts, rev(meshpts)),
        c(e_by_size_prime, rep(0, m)),
        col = adjustcolor("orange", alpha.f = 0.3), border = NA)
lines(meshpts, e_by_size_prime, lwd = 2, col = "orange")
par(mfrow = c(1, 1))
```

## Sensitivity of Lambda to Vital Rate Parameters

While the kernel-level sensitivity and elasticity are useful, we often want to know how $\lambda$ responds to changes in specific vital rate parameters. We can estimate this numerically by perturbing each parameter and recomputing $\lambda$.

```{r}
compute_lambda <- function(params) {
  dl <- list(
    surv_int = params[1], surv_z = params[2],
    grow_int = params[3], grow_z = params[4], grow_sd = params[5],
    repr_int = params[6], repr_z = params[7],
    recr_pr  = 1 / (1 + exp(-data_list$recr_int)),
    rcsz_int = params[8], rcsz_z = params[9], rcsz_sd = params[10]
  )

  ipm <- init_ipm(sim_gen = "simple", di_dd = "di", det_stoch = "det") %>%
    define_kernel(
      name = "P", formula = s * g, family = "CC",
      s = 1 / (1 + exp(-(surv_int + surv_z * z_1))),
      g = dnorm(z_2, grow_mu, grow_sd),
      grow_mu = grow_int + grow_z * z_1,
      data_list = dl, states = list(c("z")),
      uses_par_sets = FALSE, evict_cor = FALSE
    ) %>%
    define_kernel(
      name = "F", formula = s * pb * 0.5 * recr_pr * c_z1z, family = "CC",
      s = 1 / (1 + exp(-(surv_int + surv_z * z_1))),
      pb = 1 / (1 + exp(-(repr_int + repr_z * z_1))),
      c_z1z = dnorm(z_2, rcsz_mu, rcsz_sd),
      rcsz_mu = rcsz_int + rcsz_z * z_1,
      data_list = dl, states = list(c("z")),
      uses_par_sets = FALSE, evict_cor = FALSE
    ) %>%
    define_impl(make_impl_args_list(
      kernel_names = c("P", "F"), int_rule = rep("midpoint", 2),
      state_start = rep("z", 2), state_end = rep("z", 2)
    )) %>%
    define_domains(z = c(L, U, m)) %>%
    define_pop_state(n_z = rep(1 / m, m)) %>%
    make_ipm(iterations = 100)

  return(lambda(ipm))
}

base_params <- c(data_list$surv_int, data_list$surv_z,
                 data_list$grow_int, data_list$grow_z, data_list$grow_sd,
                 data_list$repr_int, data_list$repr_z,
                 data_list$rcsz_int, data_list$rcsz_z, data_list$rcsz_sd)
param_names <- c("surv_int", "surv_z", "grow_int", "grow_z", "grow_sd",
                 "repr_int", "repr_z", "rcsz_int", "rcsz_z", "rcsz_sd")

eps <- 1e-5
sensitivities_param <- numeric(length(base_params))
for (i in seq_along(base_params)) {
  p_plus <- base_params; p_plus[i] <- p_plus[i] + eps
  p_minus <- base_params; p_minus[i] <- p_minus[i] - eps
  sensitivities_param[i] <- (compute_lambda(p_plus) - compute_lambda(p_minus)) / (2 * eps)
}

elasticities_param <- sensitivities_param * base_params / lam
```

```{r}
#| echo: false
cat("Parameter sensitivities and elasticities:\n")
cat(strrep("-", 55), "\n")
for (i in seq_along(param_names)) {
  cat(sprintf("  %-10s: sens = %10s, elas = %10s\n",
              param_names[i],
              formatC(sensitivities_param[i], format = "g", digits = 4),
              formatC(elasticities_param[i], format = "g", digits = 4)))
}
```

```{r}
barplot(abs(elasticities_param), names.arg = param_names,
        ylab = "|Elasticity|",
        main = expression("Absolute elasticity of " * lambda * " to vital rate parameters"),
        las = 2, cex.names = 0.8)
```

## Summary

In this vignette we:

1. Computed the **sensitivity matrix** $S(z', z) = v(z') w(z) / \langle v, w \rangle$
2. Computed the **elasticity matrix** $e(z', z) = K(z', z) S(z', z) / \lambda$
3. Verified that elasticities sum to 1
4. Decomposed elasticities into P and F contributions — for Soay sheep, survival-growth (P) dominates
5. Computed numerical sensitivities and elasticities of $\lambda$ to individual vital rate parameters
