---
title: "Ungulate IPM: Soay Sheep"
author: "Simon Frost"
format:
  html:
    code-fold: false
    toc: true
---

## Overview

This vignette builds a density-independent, deterministic IPM for **Soay sheep** (*Ovis aries*), an iteroparous ungulate from the St Kilda archipelago, Scotland. Unlike the monocarpic *Oenothera* in the previous vignette, Soay sheep can reproduce multiple times — survival appears in *both* the P and F kernels.

Key differences from the monocarp model:

- Survival is in both P and F kernels (reproductive individuals survive)
- A 0.5 sex ratio factor in the F kernel (females-only model)
- Recruit size depends on maternal size (not fixed)
- Recruitment probability is size-independent

## Setup

```{r}
library(ipmr)
```

## Parameters

Size is log body mass. Parameters are from the textbook (Ellner, Childs & Rees, Chapter 2).

```{r}
data_list <- list(
  surv_int  = -9.65,
  surv_z    =  3.77,
  grow_int  =  1.41,
  grow_z    =  0.557,
  grow_sd   =  0.0799,
  repr_int  = -7.23,
  repr_z    =  2.60,
  recr_int  =  1.93,
  recr_pr   =  1 / (1 + exp(-1.93)),  # pre-computed recruitment probability
  rcsz_int  =  0.362,
  rcsz_z    =  0.709,
  rcsz_sd   =  0.159
)
```

## Domain

```{r}
L <- 1.6
U <- 3.7
m <- 100
```

## Kernel Structure

For an iteroparous species, the kernel decomposition is:

$$
P(z', z) = s(z) \cdot g(z'|z)
$$

$$
F(z', z) = s(z) \cdot p_b(z) \cdot \frac{1}{2} \cdot p_r \cdot c(z'|z)
$$

Note that **survival $s(z)$ appears in F** because a mother must survive to census to be counted with her offspring. The factor $\frac{1}{2}$ accounts for the sex ratio (we model females only).

```{r}
ungulate_ipm <- init_ipm(sim_gen = "simple", di_dd = "di", det_stoch = "det") %>%
  define_kernel(
    name          = "P",
    formula       = s * g,
    family        = "CC",
    s             = 1 / (1 + exp(-(surv_int + surv_z * z_1))),
    g             = dnorm(z_2, grow_mu, grow_sd),
    grow_mu       = grow_int + grow_z * z_1,
    data_list     = data_list,
    states        = list(c("z")),
    uses_par_sets = FALSE,
    evict_cor     = FALSE
  ) %>%
  define_kernel(
    name          = "F",
    formula       = s * pb * 0.5 * recr_pr * c_z1z,
    family        = "CC",
    s             = 1 / (1 + exp(-(surv_int + surv_z * z_1))),
    pb            = 1 / (1 + exp(-(repr_int + repr_z * z_1))),
    c_z1z         = dnorm(z_2, rcsz_mu, rcsz_sd),
    rcsz_mu       = rcsz_int + rcsz_z * z_1,
    data_list     = data_list,
    states        = list(c("z")),
    uses_par_sets = FALSE,
    evict_cor     = FALSE
  ) %>%
  define_impl(
    make_impl_args_list(
      kernel_names = c("P", "F"),
      int_rule     = rep("midpoint", 2),
      state_start  = rep("z", 2),
      state_end    = rep("z", 2)
    )
  ) %>%
  define_domains(z = c(L, U, m)) %>%
  define_pop_state(n_z = rep(1 / m, m)) %>%
  make_ipm(iterations = 100)
```

## Eigenanalysis

```{r}
lam <- lambda(ungulate_ipm)
```

```{r}
#| echo: false
cat("Asymptotic growth rate (lambda):", lam, "\n")
```

## Vital Rate Visualization

```{r}
h <- (U - L) / m
meshpts <- L + ((1:m) - 1/2) * h

s_z <- function(z) 1 / (1 + exp(-(data_list$surv_int + data_list$surv_z * z)))
pb_z <- function(z) 1 / (1 + exp(-(data_list$repr_int + data_list$repr_z * z)))

par(mfrow = c(2, 2))
curve(s_z, L, U, lwd = 2, xlab = "Log body mass", ylab = "Probability", main = "Survival")
curve(pb_z, L, U, lwd = 2, col = "orange", xlab = "Log body mass", ylab = "Probability",
      main = "Reproduction probability")
curve(dnorm(x, data_list$grow_int + data_list$grow_z * x, data_list$grow_sd), L, U,
      lwd = 2, col = "green4", xlab = "Log body mass", ylab = "Density",
      main = "Growth (staying same size)")
curve(data_list$rcsz_int + data_list$rcsz_z * x, L, U,
      lwd = 2, col = "purple", xlab = "Maternal log mass", ylab = "Mean offspring log mass",
      main = "Recruit size (mean)")
par(mfrow = c(1, 1))
```

## Kernel Visualization

```{r}
par(mfrow = c(1, 2))
image(meshpts, meshpts, t(ungulate_ipm$sub_kernels$P),
      xlab = "Size (t)", ylab = "Size (t+1)",
      main = "P kernel", col = hcl.colors(50, "viridis"))
image(meshpts, meshpts, t(ungulate_ipm$sub_kernels$F),
      xlab = "Size (t)", ylab = "Size (t+1)",
      main = "F kernel", col = hcl.colors(50, "viridis"))
par(mfrow = c(1, 1))
```

## Stable Distribution and Reproductive Value

```{r}
w <- right_ev(ungulate_ipm)
v <- left_ev(ungulate_ipm)
stable_dist <- w$z_w / sum(w$z_w)
repro_val <- v$z_v

par(mfrow = c(1, 2))
plot(meshpts, stable_dist, type = "l", lwd = 2, col = "blue",
     xlab = "Log body mass", ylab = "Proportion",
     main = "Stable size distribution")
polygon(c(meshpts, rev(meshpts)),
        c(stable_dist, rep(0, m)),
        col = adjustcolor("blue", alpha.f = 0.3), border = NA)
plot(meshpts, repro_val, type = "l", lwd = 2, col = "red",
     xlab = "Log body mass", ylab = "Reproductive value",
     main = "Reproductive value")
par(mfrow = c(1, 1))
```

## Mean Size at Stable Distribution

```{r}
mean_z <- sum(stable_dist * meshpts)
var_z <- sum(stable_dist * meshpts^2) - mean_z^2
```

```{r}
#| echo: false
cat("Mean log body mass:", round(mean_z, 4), "\n")
cat("SD log body mass:", round(sqrt(var_z), 4), "\n")
cat("Mean body mass (kg):", round(exp(mean_z), 2), "\n")
```

## Age-Specific Size Distributions

Even in a simple (non-age-structured) IPM, we can recover age-specific size distributions from the P and F kernels.

```{r}
P_mat <- ungulate_ipm$sub_kernels$P
F_mat <- ungulate_ipm$sub_kernels$F
n_ages <- 8

age_dists <- list()
age_dists[[1]] <- F_mat %*% stable_dist / lam
for (a in 2:n_ages) {
  age_dists[[a]] <- P_mat %*% age_dists[[a - 1]] / lam
}

mean_by_age <- sapply(1:n_ages, function(a) {
  d <- age_dists[[a]] / sum(age_dists[[a]])
  sum(d * meshpts)
})
```

```{r}
#| echo: false
cat("Mean log body mass by age:\n")
for (a in 1:n_ages) {
  cat(sprintf("  Age %d: %.3f (mass = %.2f kg)\n",
              a - 1, mean_by_age[a], exp(mean_by_age[a])))
}
```

```{r}
cols <- rainbow(min(n_ages, 6))
plot(meshpts, age_dists[[1]] / sum(age_dists[[1]]),
     type = "l", lwd = 2, col = cols[1],
     xlab = "Log body mass", ylab = "Density",
     main = "Age-specific size distributions",
     ylim = c(0, max(sapply(1:6, function(a) max(age_dists[[a]] / sum(age_dists[[a]]))))))
for (a in 2:min(n_ages, 6)) {
  lines(meshpts, age_dists[[a]] / sum(age_dists[[a]]), lwd = 2, col = cols[a])
}
legend("topright", paste("Age", 0:5), col = cols, lwd = 2, cex = 0.8)
```

## Comparison with Monocarp Structure

| Feature | Monocarp (*Oenothera*) | Ungulate (Soay sheep) |
|---------|------------------------|-----------------------|
| Life history | Monocarpic (reproduce once, die) | Iteroparous (repeated reproduction) |
| Survival in F | No (flowering = death) | Yes (mothers must survive) |
| Sex ratio | Not modeled | 0.5 (females only) |
| Recruit size | Fixed distribution | Depends on maternal size |
| P kernel | $s(z)(1-p_b(z)) \cdot g(z'|z)$ | $s(z) \cdot g(z'|z)$ |
| F kernel | $p_b(z) \cdot b(z) \cdot p_r \cdot c_0(z')$ | $s(z) \cdot p_b(z) \cdot 0.5 \cdot p_r \cdot c(z'|z)$ |

## Population Projection

```{r}
all_lambdas <- lambda(ungulate_ipm, type_lambda = "all")

plot(all_lambdas, type = "l", lwd = 2,
     xlab = "Time step", ylab = expression(lambda(t)),
     main = expression("Convergence to asymptotic " * lambda))
abline(h = lam, col = "red", lty = 2, lwd = 2)
```

## Summary

In this vignette we:

1. Built a density-independent, deterministic IPM for Soay sheep using `ipmr`
2. Highlighted the key structural difference from monocarpic models: survival in the F kernel
3. Computed $\lambda$, stable distribution, and reproductive value
4. Recovered age-specific size distributions from the simple IPM
5. Compared the ungulate and monocarp model structures
