---
title: "Age x Size IPM"
author: "Simon Frost"
format:
  html:
    code-fold: false
    toc: true
---

## Overview

In many populations, vital rates depend on both **size** and **age**. An age×size IPM tracks the joint distribution $n_a(z, t)$ — the density of individuals of age $a$ and size $z$ at time $t$.

The model is implemented as a block-structured matrix where:

- **P kernels** (survival-growth) appear on the sub-diagonal: age $a$ survivors move to age $a+1$
- **F kernels** (fecundity) appear in the first row: offspring from all ages enter age class 1
- The maximum age class is absorbing: survivors at max age stay at max age

This vignette models Soay sheep with age-dependent survival, growth, reproduction, and recruitment, following Ellner, Childs & Rees (Chapter 6).

## Setup

```{r}
library(ipmr)
```

## Parameters

Age effects are linear on the logit (survival, reproduction, recruitment) or linear (growth) scale. The `_a` parameters give the per-year age effect.

```{r}
data_list <- list(
  # Survival (logistic: intercept + slope*z + age_effect*a)
  surv_int  = -17.0,
  surv_z    =  6.68,
  surv_a    = -0.334,     # strong negative: senescence

  # Growth (linear: intercept + slope*z + age_effect*a)
  grow_int  =  1.27,
  grow_z    =  0.612,
  grow_a    = -0.00724,   # slight negative: growth slows with age
  grow_sd   =  0.0787,

  # Reproduction probability (logistic)
  repr_int  = -7.88,
  repr_z    =  3.11,
  repr_a    = -0.078,     # declines with age

  # Recruitment probability (logistic, no size effect)
  recr_int  =  1.11,
  recr_a    =  0.184,     # increases with age (experienced mothers)

  # Recruit size (depends on maternal size only, NOT age)
  rcsz_int  =  0.362,
  rcsz_z    =  0.709,
  rcsz_sd   =  0.159
)
```

## Domain and Age Structure

```{r}
L <- 1.6
U <- 3.7
m <- 100
max_age <- 10

h <- (U - L) / m
meshpts <- L + ((1:m) - 1/2) * h
```

```{r}
#| echo: false
cat("Size mesh:", m, "points on [", L, ",", U, "]\n")
cat("Age classes: 0 to", max_age, "(age", max_age, "is absorbing)\n")
cat("Total state space:", m * (max_age + 1), "=", m, "×", max_age + 1, "\n")
```

## Age-Dependent Vital Rates

```{r}
s_z <- function(z, a) {
  1 / (1 + exp(-(data_list$surv_int + data_list$surv_z * z + data_list$surv_a * a)))
}

g_z1z <- function(z_prime, z, a) {
  mu <- data_list$grow_int + data_list$grow_z * z + data_list$grow_a * a
  dnorm(z_prime, mu, data_list$grow_sd)
}

pb_z <- function(z, a) {
  if (a <= 0) return(0)  # age 0 = newborns, cannot reproduce
  1 / (1 + exp(-(data_list$repr_int + data_list$repr_z * z + data_list$repr_a * a)))
}

pr_z <- function(a) {
  1 / (1 + exp(-(data_list$recr_int + data_list$recr_a * a)))
}

c_z1z <- function(z_prime, z) {
  dnorm(z_prime, data_list$rcsz_int + data_list$rcsz_z * z, data_list$rcsz_sd)
}
```

### Visualize Age Effects

```{r}
ages_plot <- c(1, 3, 5, 7, 10)
z_plot <- seq(L, U, length.out = 100)
cols <- rainbow(length(ages_plot))

par(mfrow = c(2, 2))

# Survival by age
plot(NULL, xlim = c(L, U), ylim = c(0, 1),
     xlab = "Size", ylab = "s(z)", main = "Survival by age")
for (i in seq_along(ages_plot)) {
  lines(z_plot, sapply(z_plot, s_z, a = ages_plot[i]), lwd = 2, col = cols[i])
}
legend("topleft", paste("Age", ages_plot), col = cols, lwd = 2, cex = 0.7)

# Reproduction probability by age
plot(NULL, xlim = c(L, U), ylim = c(0, 1),
     xlab = "Size", ylab = expression(p[b](z)), main = "Reproduction prob by age")
for (i in seq_along(ages_plot)) {
  lines(z_plot, sapply(z_plot, pb_z, a = ages_plot[i]), lwd = 2, col = cols[i])
}
legend("topleft", paste("Age", ages_plot), col = cols, lwd = 2, cex = 0.7)

# Recruitment probability by age
plot(1:max_age, sapply(1:max_age, pr_z), type = "b", pch = 16, lwd = 2,
     xlab = "Age", ylab = expression(p[r](a)), main = "Recruitment prob by age")

# Growth mean by age
plot(NULL, xlim = c(L, U), ylim = c(L, U),
     xlab = "Size (t)", ylab = "Mean size (t+1)", main = "Growth mean by age")
for (i in seq_along(ages_plot)) {
  a <- ages_plot[i]
  lines(z_plot, data_list$grow_int + data_list$grow_z * z_plot + data_list$grow_a * a,
        lwd = 2, col = cols[i])
}
legend("topleft", paste("Age", ages_plot), col = cols, lwd = 2, cex = 0.7)
par(mfrow = c(1, 1))
```

## Building Age-Specific Kernels (Manual Construction)

We build the full block-structured matrix manually, constructing age-specific P and F kernel matrices.

```{r}
# Build P kernel matrix for a given age
build_P <- function(a) {
  P <- matrix(0, m, m)
  for (j in 1:m) {
    for (i in 1:m) {
      P[i, j] <- s_z(meshpts[j], a) * g_z1z(meshpts[i], meshpts[j], a) * h
    }
  }
  return(P)
}

# Build F kernel matrix for a given age
build_F <- function(a) {
  F_mat <- matrix(0, m, m)
  for (j in 1:m) {
    for (i in 1:m) {
      F_mat[i, j] <- s_z(meshpts[j], a) * pb_z(meshpts[j], a) * 0.5 *
        pr_z(a) * c_z1z(meshpts[i], meshpts[j]) * h
    }
  }
  return(F_mat)
}
```

```{r}
# Assemble the full block matrix
# Ages 0, 1, ..., max_age => max_age + 1 age classes
n_ages <- max_age + 1
total_dim <- m * n_ages

K_age <- matrix(0, total_dim, total_dim)

for (a in 0:max_age) {
  # Row/column blocks for age a
  r_a <- (a * m + 1):((a + 1) * m)

  # F kernel: all ages contribute offspring to age-0 block (first row of blocks)
  F_a <- build_F(a)
  K_age[1:m, r_a] <- F_a

  if (a < max_age) {
    # P kernel: age a survivors move to age a+1 (sub-diagonal)
    P_a <- build_P(a)
    r_next <- ((a + 1) * m + 1):((a + 2) * m)
    K_age[r_next, r_a] <- P_a
  } else {
    # Max age is absorbing: survivors stay at max age
    P_max <- build_P(max_age)
    K_age[r_a, r_a] <- P_max
  }
}
```

```{r}
#| echo: false
cat("Full kernel size:", nrow(K_age), "x", ncol(K_age), "\n")
cat("  =", n_ages, "age classes ×", m, "size classes\n")
```

## Eigenanalysis

```{r}
e_age <- eigen(K_age)
lam_age <- Re(e_age$values[1])
```

```{r}
#| echo: false
cat("Age×size model λ =", round(lam_age, 6), "\n")
```

### Comparison with Size-Only Model

The size-only model (no age structure) gives a different $\lambda$ because it does not account for age-specific vital rate variation.

```{r}
s_z_avg <- function(z) mean(sapply(0:max_age, function(a) s_z(z, a)))
g_z1z_avg <- function(z_prime, z) mean(sapply(0:max_age, function(a) g_z1z(z_prime, z, a)))
fec_avg <- function(z_prime, z) {
  mean(sapply(0:max_age, function(a) {
    s_z(z, a) * pb_z(z, a) * 0.5 * pr_z(a) * c_z1z(z_prime, z)
  }))
}

K_avg <- matrix(0, m, m)
for (j in 1:m) {
  for (i in 1:m) {
    K_avg[i, j] <- (s_z_avg(meshpts[j]) * g_z1z_avg(meshpts[i], meshpts[j]) +
                       fec_avg(meshpts[i], meshpts[j])) * h
  }
}

e_avg <- eigen(K_avg)
lam_avg <- Re(e_avg$values[1])
```

```{r}
#| echo: false
cat("Size-only (age-averaged) λ =", round(lam_avg, 6), "\n")
cat("Age×size λ                 =", round(lam_age, 6), "\n")
cat("Difference                 =", round(lam_age - lam_avg, 6), "\n")
```

The models give different $\lambda$ values because the age-averaged model does not properly account for the covariance between age and size in the stable population.

## Block Matrix Structure

```{r}
image(1:total_dim, 1:total_dim, log10(abs(t(K_age)) + 1e-20),
      xlab = "State index", ylab = "State index",
      main = paste0("log₁₀|K| (", n_ages, " ages × ", m, " sizes)"),
      col = hcl.colors(50, "viridis"))

# Draw block boundaries
for (a in 1:max_age) {
  boundary <- a * m + 0.5
  abline(v = boundary, col = "white", lwd = 0.5)
  abline(h = boundary, col = "white", lwd = 0.5)
}
```

The block structure shows:

- **First row block**: F kernels (fecundity from all ages producing age-0 offspring)
- **Sub-diagonal blocks**: P kernels (survival-growth, aging from $a$ to $a+1$)
- **Last diagonal block**: P kernel for max age (absorbing: survivors stay at max age)

## Stable Age-Size Distribution

```{r}
w_age <- Re(e_age$vectors[, 1])
w_age <- w_age / sum(w_age)

# Extract age-specific distributions
age_dists <- list()
age_totals <- numeric(n_ages)
for (a in 0:max_age) {
  idx <- (a * m + 1):((a + 1) * m)
  age_dists[[a + 1]] <- w_age[idx]
  age_totals[a + 1] <- sum(w_age[idx])
}
```

```{r}
#| echo: false
cat("Stable age distribution:\n")
for (a in 0:max_age) {
  cat(sprintf("  Age %d: %.2f%%\n", a, age_totals[a + 1] * 100))
}
```

```{r}
barplot(age_totals, names.arg = 0:max_age,
        xlab = "Age class", ylab = "Proportion",
        main = "Stable age distribution")
```

```{r}
# Age-specific size distributions (normalized within each age)
cols <- rainbow(min(n_ages, 6))

max_y <- 0
for (a in 0:min(max_age, 5)) {
  if (sum(age_dists[[a + 1]]) > 1e-10) {
    d <- age_dists[[a + 1]] / sum(age_dists[[a + 1]])
    max_y <- max(max_y, max(d))
  }
}
plot(NULL, xlim = c(L, U), ylim = c(0, max_y * 1.1),
     xlab = "Log body mass", ylab = "Density",
     main = "Stable size distribution by age")
for (a in 0:min(max_age, 5)) {
  if (sum(age_dists[[a + 1]]) > 1e-10) {
    d <- age_dists[[a + 1]] / sum(age_dists[[a + 1]])
    lines(meshpts, d, lwd = 2, col = cols[a + 1])
  }
}
legend("topright", paste("Age", 0:5), col = cols[1:6], lwd = 2, cex = 0.8)
```

```{r}
# Mean size by age
mean_by_age <- numeric(n_ages)
for (a in 0:max_age) {
  if (sum(age_dists[[a + 1]]) > 1e-10) {
    d <- age_dists[[a + 1]] / sum(age_dists[[a + 1]])
    mean_by_age[a + 1] <- sum(d * meshpts)
  } else {
    mean_by_age[a + 1] <- NA
  }
}

plot(0:max_age, mean_by_age, type = "b", pch = 16, lwd = 2,
     xlab = "Age", ylab = "Mean log body mass",
     main = "Mean size by age at stable distribution")
```

## Population Dynamics

```{r}
# Initial population: all in age class 0 (newborns)
n0_age <- rep(0, total_dim)
n0_age[1:m] <- 1 / m

# Iterate
n_steps <- 100
n_t <- n0_age
lambdas_age <- numeric(n_steps)

for (t in 1:n_steps) {
  n_next <- K_age %*% n_t
  lambdas_age[t] <- sum(n_next) / sum(n_t)
  n_t <- n_next
}

plot(lambdas_age, type = "l", lwd = 1,
     xlab = "Time step", ylab = expression(lambda(t)),
     main = expression("Convergence to asymptotic " * lambda * " (age×size model)"))
abline(h = lam_age, col = "red", lty = 2, lwd = 2)
legend("topright", c(expression(lambda(t)), expression("Asymptotic " * lambda)),
       lty = c(1, 2), col = c("black", "red"), lwd = 2)
```

## Age-Specific Contributions

```{r}
# Which age classes contribute most to population growth?
w_r <- Re(e_age$vectors[, 1])
w_r <- w_r / sum(w_r)

e_left <- eigen(t(K_age))
v <- Re(e_left$vectors[, 1])
v <- v / v[1]

# Sensitivity matrix
S_age <- outer(v, w_r) / sum(v * w_r)

# Fecundity elasticity by age
fec_elas_by_age <- numeric(n_ages)
for (a in 0:max_age) {
  c_range <- (a * m + 1):((a + 1) * m)
  r_range <- 1:m
  E_Fa <- (K_age[r_range, c_range] * S_age[r_range, c_range]) / lam_age
  fec_elas_by_age[a + 1] <- sum(E_Fa)
}

barplot(fec_elas_by_age, names.arg = 0:max_age,
        xlab = "Age", ylab = "Fecundity elasticity",
        main = "Fecundity elasticity by age class")
```

## Summary

In this vignette we:

1. Built an **age×size IPM** for Soay sheep with age-dependent vital rates
2. Constructed the full block matrix with age-specific P and F kernels
3. Visualized age effects on survival (senescence), growth (slowing), and reproduction (decline)
4. Computed $\lambda$ and compared with the size-only model
5. Examined the stable age-size distribution showing how size distributions shift with age
6. Computed age-specific fecundity elasticities
