---
title: "Stochastic IPMs"
author: "Simon Frost"
format:
  html:
    code-fold: false
    toc: true
---

## Overview

In stochastic IPMs, the environment varies over time, causing the projection kernel to change at each time step. There are two main approaches:

1. **Kernel-resampled**: pre-compute a set of kernels (e.g., one per observed year) and randomly sample from them each time step
2. **Parameter-resampled**: sample environmental parameters each time step and build a new kernel

The **stochastic growth rate** is the geometric mean of per-step growth rates:

$$
\log \lambda_s = \lim_{T \to \infty} \frac{1}{T} \sum_{t=1}^{T} \log \lambda_t
$$

A key result is that $\lambda_s \leq \lambda_{\text{det}}$ — environmental stochasticity always reduces long-term growth relative to the deterministic case (Tuljapurkar's inequality).

## Setup

```{r}
library(ipmr)
```

## Part 1: Kernel-Resampled Stochastic IPM

We build 5 year-specific kernels for a monocarpic perennial, each with different random intercepts for survival, growth, and fecundity. This follows the `ipmr` test case.

### Parameters

```{r}
s_int    <- 1.03;    s_slope  <- 2.2
g_int    <- 8.0;     g_slope  <- 0.92;  sd_g <- 0.9
f_r_int  <- 0.09;    f_r_slope <- 0.05
f_s_int  <- 0.1;     f_s_slope <- 0.005
mu_fd    <- 9.0;     sd_fd    <- 2.0

# Year-specific random intercepts
set.seed(50127)
g_r   <- rnorm(5) * 0.3     # growth random intercepts
s_r   <- rnorm(5) * 0.7     # survival random intercepts
f_s_r <- rnorm(5) * 0.2     # fecundity random intercepts

# Domain
L_stoch <- 0.2;  U_stoch <- 40.0;  n_stoch <- 100
```

### Build Year-Specific Kernels with ipmr

The `ipmr` package supports kernel-resampled stochastic IPMs using `uses_par_sets = TRUE` and `par_set_indices`. Each year gets its own kernel with year-specific random intercepts.

```{r}
# Prepare data list with year-specific parameters
params <- list(
  s_int     = s_int,
  s_slope   = s_slope,
  g_int     = g_int,
  g_slope   = g_slope,
  sd_g      = sd_g,
  f_r_int   = f_r_int,
  f_r_slope = f_r_slope,
  f_s_int   = f_s_int,
  f_s_slope = f_s_slope,
  mu_fd     = mu_fd,
  sd_fd     = sd_fd
)

# Add year-specific random intercepts: g_r_1, g_r_2, ..., s_r_1, ..., f_s_r_1, ...
for (i in 1:5) {
  params[[paste0("g_r_", i)]]   <- g_r[i]
  params[[paste0("s_r_", i)]]   <- s_r[i]
  params[[paste0("f_s_r_", i)]] <- f_s_r[i]
}

par_set_indices <- list(yr = 1:5)
```

```{r}
stoch_kern_ipm <- init_ipm(sim_gen = "simple", di_dd = "di",
                           det_stoch = "stoch", kern_param = "kern") %>%
  define_kernel(
    name            = "P_yr",
    formula         = s_yr * g_yr,
    family          = "CC",
    s_yr            = 1 / (1 + exp(-(s_int + s_slope * ht_1 + s_r_yr))),
    g_yr            = dnorm(ht_2, mu_g_yr, sd_g),
    mu_g_yr         = g_int + g_slope * ht_1 + g_r_yr,
    data_list       = params,
    states          = list(c("ht")),
    uses_par_sets   = TRUE,
    par_set_indices = par_set_indices,
    evict_cor       = TRUE,
    evict_fun       = truncated_distributions("norm", "g_yr")
  ) %>%
  define_kernel(
    name            = "F_yr",
    formula         = f_r * f_s_yr * f_d,
    family          = "CC",
    f_r             = 1 / (1 + exp(-(f_r_int + f_r_slope * ht_1))),
    f_s_yr          = exp(f_s_int + f_s_slope * ht_1 + f_s_r_yr),
    f_d             = dnorm(ht_2, mu_fd, sd_fd),
    data_list       = params,
    states          = list(c("ht")),
    uses_par_sets   = TRUE,
    par_set_indices = par_set_indices,
    evict_cor       = TRUE,
    evict_fun       = truncated_distributions("norm", "f_d")
  ) %>%
  define_impl(
    make_impl_args_list(
      kernel_names = c("P_yr", "F_yr"),
      int_rule     = rep("midpoint", 2),
      state_start  = rep("ht", 2),
      state_end    = rep("ht", 2)
    )
  ) %>%
  define_domains(ht = c(L_stoch, U_stoch, n_stoch)) %>%
  define_pop_state(n_ht = runif(n_stoch))

# Deterministic lambda for each year
for (yr in 1:5) {
  temp_ipm <- make_ipm(stoch_kern_ipm,
                       iterate = TRUE, iterations = 100,
                       normalize_pop_size = TRUE,
                       kernel_seq = rep(yr, 100))
  cat("Year", yr, ": λ =", round(lambda(temp_ipm), 4), "\n")
}
```

### Deterministic Lambda (Average Kernel)

```{r}
# Run with cycling through all years deterministically
det_ipm <- make_ipm(stoch_kern_ipm,
                    iterate = TRUE, iterations = 500,
                    normalize_pop_size = TRUE,
                    kernel_seq = rep(1:5, 100))
lam_det <- lambda(det_ipm)
```

```{r}
#| echo: false
cat("Deterministic λ (cycling kernel):", round(lam_det, 6), "\n")
```

### Stochastic Simulation

```{r}
set.seed(42)
n_iter <- 1000
kern_seq <- sample(1:5, n_iter, replace = TRUE)

stoch_ipm <- make_ipm(stoch_kern_ipm,
                      iterate = TRUE, iterations = n_iter,
                      normalize_pop_size = TRUE,
                      kernel_seq = kern_seq)
```

### Stochastic Growth Rate

```{r}
all_lambdas <- lambda(stoch_ipm, type_lambda = "all")

# Stochastic growth rate: exp(mean(log(lambda_t)))
burn_in <- 100
log_lambdas <- log(all_lambdas[(burn_in + 1):length(all_lambdas)])
log_lam_s <- mean(log_lambdas)
lam_s <- exp(log_lam_s)
```

```{r}
#| echo: false
cat("Stochastic growth rate:     λ_s =", round(lam_s, 6), "\n")
cat("Log stochastic growth rate: log(λ_s) =", round(log_lam_s, 6), "\n")
cat("Deterministic growth rate:  λ_det =", round(lam_det, 6), "\n")
cat("\nTuljapurkar's inequality: λ_s ≤ λ_det?", lam_s <= lam_det, "\n")
```

### Visualize Stochastic Dynamics

```{r}
plot(all_lambdas[1:200], type = "l", lwd = 0.5, col = adjustcolor("black", 0.7),
     xlab = "Time step", ylab = expression(lambda(t)),
     main = "Per-step growth rates (kernel-resampled)")
abline(h = lam_s, col = "red", lwd = 2)
abline(h = lam_det, col = "blue", lwd = 2, lty = 2)
legend("topright",
       c(expression(lambda[s] ~ "(stochastic)"), expression(lambda[det] ~ "(deterministic)")),
       col = c("red", "blue"), lwd = 2, lty = c(1, 2))
```

```{r}
# Running average of log lambda
log_all <- log(all_lambdas)
running_mean <- cumsum(log_all[(burn_in + 1):length(log_all)]) /
  seq_along(log_all[(burn_in + 1):length(log_all)])

plot((burn_in + 1):length(log_all), running_mean, type = "l", lwd = 1,
     xlab = "Time step", ylab = "Running mean log(λ)",
     main = "Convergence of stochastic growth rate")
abline(h = log_lam_s, col = "red", lwd = 2)
legend("topright", expression("log(" * lambda[s] * ")"), col = "red", lwd = 2)
```

## Part 2: Parameter-Resampled Stochastic IPM

Instead of pre-computing kernels, we sample parameters from environmental distributions at each time step. In `ipmr`, this uses `define_env_state()`.

```{r}
stoch_param_ipm <- init_ipm(sim_gen = "simple", di_dd = "di",
                            det_stoch = "stoch", kern_param = "param") %>%
  define_kernel(
    name      = "P",
    formula   = s * g,
    family    = "CC",
    s         = 1 / (1 + exp(-(s_int + s_slope * ht_1 + s_offset))),
    g         = dnorm(ht_2, mu_g, sd_g),
    mu_g      = g_int + g_slope * ht_1 + g_offset,
    data_list = list(s_int = s_int, s_slope = s_slope,
                     g_int = g_int, g_slope = g_slope, sd_g = sd_g),
    states    = list(c("ht")),
    evict_cor = TRUE,
    evict_fun = truncated_distributions("norm", "g")
  ) %>%
  define_kernel(
    name      = "F",
    formula   = f_r * f_s * f_d,
    family    = "CC",
    f_r       = 1 / (1 + exp(-(f_r_int + f_r_slope * ht_1))),
    f_s       = exp(f_s_int + f_s_slope * ht_1 + f_s_offset),
    f_d       = dnorm(ht_2, mu_fd, sd_fd),
    data_list = list(f_r_int = f_r_int, f_r_slope = f_r_slope,
                     f_s_int = f_s_int, f_s_slope = f_s_slope,
                     mu_fd = mu_fd, sd_fd = sd_fd),
    states    = list(c("ht")),
    evict_cor = TRUE,
    evict_fun = truncated_distributions("norm", "f_d")
  ) %>%
  define_impl(
    make_impl_args_list(
      kernel_names = c("P", "F"),
      int_rule     = rep("midpoint", 2),
      state_start  = rep("ht", 2),
      state_end    = rep("ht", 2)
    )
  ) %>%
  define_domains(ht = c(L_stoch, U_stoch, n_stoch)) %>%
  define_env_state(
    s_offset   = rnorm(1, 0, 0.7),
    g_offset   = rnorm(1, 0, 0.3),
    f_s_offset = rnorm(1, 0, 0.2),
    data_list  = list()
  ) %>%
  define_pop_state(n_ht = runif(n_stoch)) %>%
  make_ipm(iterate = TRUE, iterations = 1000,
           normalize_pop_size = TRUE)
```

```{r}
all_lambdas_param <- lambda(stoch_param_ipm, type_lambda = "all")

log_lambdas_param <- log(all_lambdas_param[(burn_in + 1):length(all_lambdas_param)])
lam_s_param <- exp(mean(log_lambdas_param))
```

```{r}
#| echo: false
cat("Stochastic growth rate (parameter-resampled):", round(lam_s_param, 6), "\n")
```

```{r}
plot(all_lambdas_param[1:200], type = "l", lwd = 0.5, col = adjustcolor("black", 0.7),
     xlab = "Time step", ylab = expression(lambda(t)),
     main = "Per-step growth rates (parameter-resampled)")
abline(h = lam_s_param, col = "red", lwd = 2)
legend("topright", expression(lambda[s]), col = "red", lwd = 2)
```

## Comparing the Two Approaches

| Feature | Kernel-resampled | Parameter-resampled |
|---------|-----------------|---------------------|
| Pre-computation | Kernels built once | Kernel built each step |
| Environmental variation | Discrete set of environments | Continuous distribution |
| Reproducibility | Easy with fixed sequence | Requires seeded RNG |
| Speed | Faster (reuse materialized kernels) | Slower (kernel rebuilt each step) |
| Flexibility | Limited to observed environments | Can model any distribution |

## Effect of Environmental Variance

Let us examine how increasing environmental variance affects the stochastic growth rate.

```{r}
set.seed(999)
sigma_vals <- seq(0.0, 1.5, length.out = 8)
lam_s_by_sigma <- numeric(length(sigma_vals))

for (k in seq_along(sigma_vals)) {
  sigma <- sigma_vals[k]

  if (sigma < 1e-10) {
    # Deterministic case
    det_ipm_k <- init_ipm(sim_gen = "simple", di_dd = "di", det_stoch = "det") %>%
      define_kernel(
        name = "P", formula = s * g, family = "CC",
        s = 1 / (1 + exp(-(s_int + s_slope * ht_1))),
        g = dnorm(ht_2, mu_g, sd_g),
        mu_g = g_int + g_slope * ht_1,
        data_list = list(s_int = s_int, s_slope = s_slope,
                         g_int = g_int, g_slope = g_slope, sd_g = sd_g),
        states = list(c("ht")), uses_par_sets = FALSE,
        evict_cor = TRUE, evict_fun = truncated_distributions("norm", "g")
      ) %>%
      define_kernel(
        name = "F", formula = f_r * f_s * f_d, family = "CC",
        f_r = 1 / (1 + exp(-(f_r_int + f_r_slope * ht_1))),
        f_s = exp(f_s_int + f_s_slope * ht_1),
        f_d = dnorm(ht_2, mu_fd, sd_fd),
        data_list = list(f_r_int = f_r_int, f_r_slope = f_r_slope,
                         f_s_int = f_s_int, f_s_slope = f_s_slope,
                         mu_fd = mu_fd, sd_fd = sd_fd),
        states = list(c("ht")), uses_par_sets = FALSE,
        evict_cor = TRUE, evict_fun = truncated_distributions("norm", "f_d")
      ) %>%
      define_impl(make_impl_args_list(
        kernel_names = c("P", "F"), int_rule = rep("midpoint", 2),
        state_start = rep("ht", 2), state_end = rep("ht", 2)
      )) %>%
      define_domains(ht = c(L_stoch, U_stoch, n_stoch)) %>%
      define_pop_state(n_ht = runif(n_stoch)) %>%
      make_ipm(iterations = 100)
    lam_s_by_sigma[k] <- lambda(det_ipm_k)
  } else {
    stoch_k <- init_ipm(sim_gen = "simple", di_dd = "di",
                        det_stoch = "stoch", kern_param = "param") %>%
      define_kernel(
        name = "P", formula = s * g, family = "CC",
        s = 1 / (1 + exp(-(s_int + s_slope * ht_1 + s_offset))),
        g = dnorm(ht_2, mu_g, sd_g),
        mu_g = g_int + g_slope * ht_1 + g_offset,
        data_list = list(s_int = s_int, s_slope = s_slope,
                         g_int = g_int, g_slope = g_slope, sd_g = sd_g),
        states = list(c("ht")),
        evict_cor = TRUE, evict_fun = truncated_distributions("norm", "g")
      ) %>%
      define_kernel(
        name = "F", formula = f_r * f_s * f_d, family = "CC",
        f_r = 1 / (1 + exp(-(f_r_int + f_r_slope * ht_1))),
        f_s = exp(f_s_int + f_s_slope * ht_1 + f_s_offset),
        f_d = dnorm(ht_2, mu_fd, sd_fd),
        data_list = list(f_r_int = f_r_int, f_r_slope = f_r_slope,
                         f_s_int = f_s_int, f_s_slope = f_s_slope,
                         mu_fd = mu_fd, sd_fd = sd_fd),
        states = list(c("ht")),
        evict_cor = TRUE, evict_fun = truncated_distributions("norm", "f_d")
      ) %>%
      define_impl(make_impl_args_list(
        kernel_names = c("P", "F"), int_rule = rep("midpoint", 2),
        state_start = rep("ht", 2), state_end = rep("ht", 2)
      )) %>%
      define_domains(ht = c(L_stoch, U_stoch, n_stoch)) %>%
      define_env_state(
        s_offset   = rnorm(1, 0, sigma_val),
        g_offset   = rnorm(1, 0, 0.3),
        f_s_offset = rnorm(1, 0, 0.2),
        data_list  = list(sigma_val = sigma)
      ) %>%
      define_pop_state(n_ht = runif(n_stoch)) %>%
      make_ipm(iterate = TRUE, iterations = 500,
               normalize_pop_size = TRUE)

    all_lam_k <- lambda(stoch_k, type_lambda = "all")
    lam_s_by_sigma[k] <- exp(mean(log(all_lam_k[51:length(all_lam_k)])))
  }
}

plot(sigma_vals, lam_s_by_sigma, type = "b", lwd = 2, pch = 16,
     xlab = "Survival SD (σ)", ylab = expression(lambda[s]),
     main = "Stochastic growth rate vs environmental variance")
abline(h = lam_s_by_sigma[1], col = "red", lty = 2, lwd = 2)
legend("topright", expression("Deterministic " * lambda ~ "(σ=0)"),
       col = "red", lty = 2, lwd = 2)
```

As environmental variance increases, the stochastic growth rate decreases, consistent with Tuljapurkar's inequality.

## Summary

In this vignette we:

1. Built a **kernel-resampled** stochastic IPM with 5 year-specific kernels using `uses_par_sets`
2. Computed the stochastic growth rate $\lambda_s = \exp(\overline{\log \lambda_t})$
3. Verified Tuljapurkar's inequality: $\lambda_s \leq \lambda_{\text{det}}$
4. Built a **parameter-resampled** stochastic IPM using `define_env_state()`
5. Demonstrated that increasing environmental variance reduces $\lambda_s$
