---
title: "Categorical Composition of IPMs"
author: "Simon Frost"
format:
  html:
    code-fold: false
    toc: true
    toc-depth: 2
    number-sections: true
---

## Introduction

Integral Projection Models can become complex: survival, growth, fecundity, seed banks, age structure, spatial patches. **Applied category theory** provides a principled framework for composing such models from smaller, reusable parts.

This R companion vignette implements the core "lowering" functions (left/right Kan extensions, stratification, coarsening) in base R and demonstrates them on the *Ligustrum obtusifolium* model from [vignette 07](07_general_ipm.qmd). For full operadic composition via undirected wiring diagrams, see the Julia vignette which uses Catlab.jl.

## Setup

No external packages are required beyond base R.

```{r}
# No special packages needed
```

## The Ligustrum Model

### Parameters

```{r}
ctrl <- list(
  g_int     = 5.781,
  g_slope   = 0.988,
  g_sd      = 20.55699,
  s_int     = -0.352,
  s_slope   = 0.122,
  s_slope_2 = -0.000213,
  f_r_int   = -11.46,
  f_r_slope = 0.0835,
  f_s_int   = 2.6204,
  f_s_slope = 0.01256,
  f_d_mu    = 5.6655,
  f_d_sd    = 2.0734,
  e_p       = 0.15,
  g_i       = 0.5067
)

cr <- list(
  g_int     = 7.229,
  g_slope   = 0.988,
  g_sd      = 21.72262,
  s_int     = 0.0209,
  s_slope   = 0.0831,
  s_slope_2 = -0.00012999,
  f_r_int   = -11.46,
  f_r_slope = 0.0835,
  f_s_int   = 2.6204,
  f_s_slope = 0.01256,
  f_d_mu    = 5.6655,
  f_d_sd    = 2.0734,
  e_p       = 0.15,
  g_i       = 0.5067
)
```

### Domain and Vital Rates

```{r}
L <- 1.02
U <- 624.0
n <- 500

# Meshpoints and step size (midpoint rule)
h_val <- (U - L) / n
z <- L + ((1:n) - 0.5) * h_val

# Vital rate functions
survival <- function(z, p) {
  1 / (1 + exp(-(p$s_int + p$s_slope * z + p$s_slope_2 * z^2)))
}

growth <- function(z_prime, z, p) {
  mu <- p$g_int + p$g_slope * z
  ev <- pnorm(U, mu, p$g_sd) - pnorm(L, mu, p$g_sd)
  ifelse(ev > 0, dnorm(z_prime, mu, p$g_sd) / ev, dnorm(z_prime, mu, p$g_sd))
}

repr_prob <- function(z, p) {
  1 / (1 + exp(-(p$f_r_int + p$f_r_slope * z)))
}

seed_prod <- function(z, p) {
  exp(p$f_s_int + p$f_s_slope * z)
}

offspring_size <- function(z_prime, p) {
  d_mu <- p$f_d_mu
  d_sd <- p$f_d_sd
  ev <- pnorm(U, d_mu, d_sd) - pnorm(L, d_mu, d_sd)
  ifelse(ev > 0, dnorm(z_prime, d_mu, d_sd) / ev, dnorm(z_prime, d_mu, d_sd))
}
```

## Core Categorical Functions

### Left Kan Extension: IPM → MPM

The **left Kan extension** discretises a continuous kernel $K(z', z)$ into a matrix using the midpoint rule:

$$A_{ij} = h \cdot K(z_i, z_j)$$

```{r}
left_kan_extension <- function(kernel_fn, lower, upper, n_bins) {
  h <- (upper - lower) / n_bins
  midpoints <- lower + ((1:n_bins) - 0.5) * h
  A <- matrix(0, n_bins, n_bins)
  for (j in 1:n_bins) {
    for (i in 1:n_bins) {
      A[i, j] <- h * kernel_fn(midpoints[i], midpoints[j])
    }
  }
  list(A = A, midpoints = midpoints, h = h)
}
```

### Right Kan Extension: MPM → IPM

The **right Kan extension** constructs a piecewise-constant kernel from a matrix:

$$K_{\mathrm{pw}}(z', z) = \frac{A_{ij}}{h} \quad \text{where } z \in B_j,\; z' \in B_i$$

```{r}
right_kan_extension <- function(A, z_min, h) {
  n_bins <- nrow(A)
  function(z_new, z) {
    j <- pmin(pmax(ceiling((z - z_min) / h), 1), n_bins)
    i <- pmin(pmax(ceiling((z_new - z_min) / h), 1), n_bins)
    A[cbind(i, j)] / h
  }
}
```

### Stratification

**Stratification** extends a local model to multiple spatial patches via a block-structured matrix:

$$A_{\mathrm{strat}}[(p_{\mathrm{to}}, i),\, (p_{\mathrm{from}}, j)] = D[p_{\mathrm{to}}, p_{\mathrm{from}}] \cdot A_{\mathrm{local}}[i, j]$$

```{r}
stratify <- function(A_local, dispersal) {
  n_bins <- nrow(A_local)
  n_patches <- nrow(dispersal)
  n_total <- n_bins * n_patches
  A_strat <- matrix(0, n_total, n_total)
  for (p_to in 1:n_patches) {
    for (p_from in 1:n_patches) {
      rows <- ((p_to - 1) * n_bins + 1):(p_to * n_bins)
      cols <- ((p_from - 1) * n_bins + 1):(p_from * n_bins)
      A_strat[rows, cols] <- dispersal[p_to, p_from] * A_local
    }
  }
  A_strat
}
```

### Coarsening via Pushforward

**Coarsening** aggregates fine bins into coarse bins, preserving information from the finer grid:

```{r}
coarsen <- function(A, n_coarse) {
  n_fine <- nrow(A)
  stopifnot(n_fine %% n_coarse == 0)
  bins_per_coarse <- n_fine %/% n_coarse

  # Coarsening map: fine bin i → coarse bin
  coarsening_map <- ((1:n_fine) - 1) %/% bins_per_coarse + 1

  A_coarse <- matrix(0, n_coarse, n_coarse)
  for (j_fine in 1:n_fine) {
    c_j <- coarsening_map[j_fine]
    for (i_fine in 1:n_fine) {
      c_i <- coarsening_map[i_fine]
      A_coarse[c_i, c_j] <- A_coarse[c_i, c_j] + A[i_fine, j_fine] / bins_per_coarse
    }
  }
  A_coarse
}
```

### Compose Kernels

Sum sub-kernel functions additively (UWD evaluation without Catlab):

```{r}
compose_kernels <- function(sub_kernels, lower, upper, n_bins) {
  h <- (upper - lower) / n_bins
  midpoints <- lower + ((1:n_bins) - 0.5) * h
  K <- matrix(0, n_bins, n_bins)
  for (kfn in sub_kernels) {
    for (j in 1:n_bins) {
      for (i in 1:n_bins) {
        K[i, j] <- K[i, j] + h * kfn(midpoints[i], midpoints[j])
      }
    }
  }
  K
}
```

## Building the Ligustrum General IPM

```{r}
build_mega <- function(p) {
  # CC: survival-growth (500×500)
  cc <- matrix(0, n, n)
  for (j in 1:n) {
    s_j <- survival(z[j], p)
    for (i in 1:n) {
      cc[i, j] <- s_j * growth(z[i], z[j], p) * h_val
    }
  }

  # CD: continuous → discrete (1×500)
  cd <- repr_prob(z, p) * seed_prod(z, p) * p$g_i
  cd <- matrix(cd, nrow = 1, ncol = n)

  # DC: discrete → continuous (500×1)
  dc <- p$e_p * offspring_size(z, p) * h_val
  dc <- matrix(dc, nrow = n, ncol = 1)

  # DD: discrete → discrete (1×1)
  dd <- matrix(0, 1, 1)

  # Assemble mega-kernel: ht first, then b
  K <- rbind(
    cbind(cc, dc),
    cbind(cd, dd)
  )
  list(K = K, cc = cc, cd = cd, dc = dc, dd = dd)
}

ipm_ctrl <- build_mega(ctrl)
ipm_cr <- build_mega(cr)

lambda_ctrl <- Re(eigen(ipm_ctrl$K)$values[1])
lambda_cr <- Re(eigen(ipm_cr$K)$values[1])
```

```{r}
#| echo: false
cat(sprintf("Control:  λ = %.4f  (published ≈ 1.22)\n", lambda_ctrl))
cat(sprintf("CR:       λ = %.4f  (published ≈ 1.26)\n", lambda_cr))
```

## Operadic Composition (Conceptual)

In the Julia vignette, we use Catlab.jl to define an **undirected wiring diagram** (UWD) that specifies how sub-processes share trait axes:

```
@relation (z, z_new) begin
    survive_grow(z, z_new)
    reproduce(z, z_new)
end
```

This captures $K = P + F$ declaratively. In R, we achieve the same by passing a named list of sub-kernel functions to `compose_kernels`:

```{r}
P_kernel <- function(z_new, z) survival(z, ctrl) * growth(z_new, z, ctrl)
F_kernel <- function(z_new, z) {
  repr_prob(z, ctrl) * seed_prod(z, ctrl) * (1 - ctrl$g_i) *
    ctrl$e_p * offspring_size(z_new, ctrl)
}

K_simple <- compose_kernels(
  list(survive_grow = P_kernel, reproduce = F_kernel),
  L, U, n
)
lambda_simple <- Re(eigen(K_simple)$values[1])
```

```{r}
#| echo: false
cat(sprintf("Simple IPM (P + F, no seed bank) λ = %.4f\n", lambda_simple))
```

## Left Kan Extension: Convergence

```{r}
# Reference at high resolution
P_fn <- function(z_new, z) survival(z, ctrl) * growth(z_new, z, ctrl)
ref <- left_kan_extension(P_fn, L, U, 500)
lambda_ref <- Re(eigen(ref$A)$values[1])
```

```{r}
#| echo: false
cat("Convergence of λ for CC block (survival-growth):\n\n")
cat(sprintf("  %7s │ %9s │ %15s\n", "n_bins", "λ", "|Δλ| from ref"))
cat(sprintf(" %s┼%s┼%s\n", strrep("─", 8), strrep("─", 11), strrep("─", 17)))

for (nb in c(5, 10, 20, 50, 100, 200, 500)) {
  res <- left_kan_extension(P_fn, L, U, nb)
  lam <- Re(eigen(res$A)$values[1])
  delta <- abs(lam - lambda_ref)
  marker <- if (nb == 500) " ← reference" else ""
  cat(sprintf("  %5d │ %9.6f │ %14.2e%s\n", nb, lam, delta, marker))
}
```

## Right Kan Extension: Round Trip

```{r}
# Discretise at n=20, then reconstruct
res_20 <- left_kan_extension(P_fn, L, U, 20)
K_pw <- right_kan_extension(res_20$A, L, res_20$h)
```

```{r}
#| echo: false
cat("Round trip: IPM → MPM(n=20) → piecewise IPM\n\n")
cat(sprintf("  %18s │ %11s │ %11s │ %8s\n",
            "(z', z)", "K_original", "K_piecewise", "|Δ|"))
cat(sprintf(" %s┼%s┼%s┼%s\n",
            strrep("─", 19), strrep("─", 13), strrep("─", 13), strrep("─", 9)))
for (pair in list(c(50, 50), c(100, 200), c(300, 300), c(500, 400))) {
  zn <- pair[1]; zv <- pair[2]
  k_orig <- P_fn(zn, zv)
  k_pw <- K_pw(zn, zv)
  delta <- abs(k_orig - k_pw)
  cat(sprintf("  (%5.0f, %5.0f)     │ %11.6f │ %11.6f │ %8.6f\n",
              zn, zv, k_orig, k_pw, delta))
}
```

## Adjunction Errors

```{r}
adjunction_errors <- function(kernel_fn, lower, upper, nb) {
  res <- left_kan_extension(kernel_fn, lower, upper, nb)
  A <- res$A
  K_pw <- right_kan_extension(A, lower, res$h)
  res_rt <- left_kan_extension(K_pw, lower, upper, nb)
  A_rt <- res_rt$A

  # Unit error
  unit_err <- norm(A_rt - A, "F") / norm(A, "F")

  # Counit error (L2 on fine grid)
  n_test <- 200
  h_test <- (upper - lower) / n_test
  test_pts <- lower + ((1:n_test) - 0.5) * h_test
  counit_err <- 0
  for (zv in test_pts) {
    for (zn in test_pts) {
      counit_err <- counit_err + (K_pw(zn, zv) - kernel_fn(zn, zv))^2
    }
  }
  counit_err <- sqrt(counit_err * h_test^2)

  lam <- Re(eigen(A)$values[1])
  lambda_err <- abs(lam - lambda_ref)

  list(unit = unit_err, counit = counit_err, lambda_err = lambda_err)
}
```

```{r}
#| echo: false
cat(sprintf("  %7s │ %11s │ %11s │ %11s\n",
            "n_bins", "Unit (η)", "Counit (ε)", "|λ - λ_ref|"))
cat(sprintf(" %s┼%s┼%s┼%s\n",
            strrep("─", 8), strrep("─", 13), strrep("─", 13), strrep("─", 12)))
for (nb in c(5, 10, 20, 50, 100, 200)) {
  errs <- adjunction_errors(P_fn, L, U, nb)
  cat(sprintf("  %5d │ %11.2e │ %11.2e │ %10.2e\n",
              nb, errs$unit, errs$counit, errs$lambda_err))
}

cat("\nUnit η ≈ 0:  midpoint rule is self-consistent under round-trip\n")
cat("Counit ε → 0: piecewise IPM converges to original as bins refine\n")
```

## Stratification: Spatial Extension

```{r}
# Symmetric 2-patch dispersal
D_sym <- matrix(c(0.9, 0.1, 0.1, 0.9), 2, 2)
K_strat_sym <- stratify(ipm_ctrl$cc, D_sym)
lambda_strat_sym <- Re(eigen(K_strat_sym)$values[1])
```

```{r}
#| echo: false
cat("Symmetric dispersal (90% stay, 10% move):\n")
cat(sprintf("  Single-patch λ = %.6f\n", lambda_ref))
cat(sprintf("  Two-patch λ    = %.6f\n", lambda_strat_sym))
cat(sprintf("  Ratio: %.6f  (≈ 1.0 for identical patches)\n",
            lambda_strat_sym / lambda_ref))
```

```{r}
# Asymmetric dispersal: source-sink dynamics
D_asym <- matrix(c(0.95, 0.05, 0.20, 0.80), 2, 2)
K_strat_asym <- stratify(ipm_ctrl$cc, D_asym)
lambda_strat_asym <- Re(eigen(K_strat_asym)$values[1])
```

```{r}
#| echo: false
cat(sprintf("Asymmetric dispersal (source-sink):\n  Two-patch λ = %.6f\n",
            lambda_strat_asym))
```

## Coarsening via Pushforward

```{r}
# Fine: 100 bins, coarse: 20 bins
res_fine <- left_kan_extension(P_fn, L, U, 100)
lambda_fine <- Re(eigen(res_fine$A)$values[1])

A_coarsened <- coarsen(res_fine$A, 20)
lambda_coarsened <- Re(eigen(A_coarsened)$values[1])

res_direct_20 <- left_kan_extension(P_fn, L, U, 20)
lambda_direct_20 <- Re(eigen(res_direct_20$A)$values[1])
```

```{r}
#| echo: false
cat(sprintf("Fine (n=100):              λ = %.6f\n", lambda_fine))
cat(sprintf("Coarsened via Lan (n=20):  λ = %.6f\n", lambda_coarsened))
cat(sprintf("Direct at n=20:            λ = %.6f\n", lambda_direct_20))
cat(sprintf("Reference (n=500):         λ = %.6f\n", lambda_ref))
```

## Commutation: Discretise then Stratify = Stratify then Discretise

```{r}
n_comm <- 50

# Path 1: discretise first, then stratify
res_p1 <- left_kan_extension(P_fn, L, U, n_comm)
A_path1 <- stratify(res_p1$A, D_sym)

# Path 2: stratify the continuous kernel, then discretise
z_comm <- L + ((1:n_comm) - 0.5) * (U - L) / n_comm
h_comm <- (U - L) / n_comm
n_patches <- 2
n_total <- n_comm * n_patches
A_path2 <- matrix(0, n_total, n_total)
for (p_from in 1:n_patches) {
  for (p_to in 1:n_patches) {
    for (j in 1:n_comm) {
      for (i in 1:n_comm) {
        row <- (p_to - 1) * n_comm + i
        col <- (p_from - 1) * n_comm + j
        A_path2[row, col] <- h_comm * D_sym[p_to, p_from] * P_fn(z_comm[i], z_comm[j])
      }
    }
  }
}

comm_error <- norm(A_path1 - A_path2, "F") / norm(A_path1, "F")
```

```{r}
#| echo: false
cat(sprintf("Commutation: ||Path1 - Path2|| / ||Path1|| = %.2e\n", comm_error))
cat(sprintf("Commutes: %s\n", comm_error < 1e-12))
```

## Summary of Categorical Constructions

| Construction | Category theory | R function | What it does |
|---|---|---|---|
| Kernel composition | Operadic algebra | `compose_kernels` | Sum sub-kernels sharing trait axes |
| IPM → MPM | Left Kan extension | `left_kan_extension` | Discretise kernel → matrix |
| MPM → IPM | Right Kan extension | `right_kan_extension` | Matrix → piecewise kernel |
| Spatial model | Pullback (stratification) | `stratify` | Local dynamics + dispersal |
| Bin aggregation | Pushforward | `coarsen` | Fine → coarse matrix |

For full UWD specification and `FinFunction`-based coarsening, see the Julia vignette which uses Catlab.jl. The R implementations here use the same mathematical operations but without the formal categorical type system.
