---
title: "Density-Dependent IPMs"
author: "Simon Frost"
format:
  html:
    code-fold: false
    toc: true
---

## Overview

In density-dependent (DD) IPMs, vital rates depend on the current population size $N_t = \int n(z, t) \, dz$. The kernel $K(z', z; N_t)$ is rebuilt at each time step, so the population dynamics are nonlinear.

This vignette models **Soay sheep** with density-dependent survival, recruitment probability, and recruit size, following Ellner, Childs & Rees (Chapter 5). The key goal is to find the **equilibrium population size** $\bar{N}$ where $\lambda(N) = 1$.

## Setup

```{r}
library(ipmr)
```

## Parameters

Density-dependent parameters have an additional `Nt` term that reduces the vital rate as population size increases.

```{r}
data_list <- list(
  # Survival (density dependent: surv_Nt < 0)
  surv_int  =  1.06,
  surv_z    =  2.09,
  surv_Nt   = -0.018,
  # Growth (NOT density dependent)
  grow_int  =  1.41,
  grow_z    =  0.557,
  grow_sd   =  0.0799,
  # Reproduction probability (NOT density dependent)
  repr_int  = -7.23,
  repr_z    =  2.60,
  # Recruitment probability (density dependent)
  recr_int  =  4.43,
  recr_Nt   = -0.00919,
  # Recruit size (density dependent)
  rcsz_int  =  0.540,
  rcsz_z    =  0.710,
  rcsz_Nt   = -0.000642,
  rcsz_sd   =  0.159
)
```

## Domain

```{r}
L <- 0.5
U <- 3.8
m <- 165
```

## Density-Dependent Vital Rates

```{r}
s_z <- function(z, Nt) 1 / (1 + exp(-(data_list$surv_int + data_list$surv_z * z + data_list$surv_Nt * Nt)))
g_z1z <- function(z_prime, z) dnorm(z_prime, data_list$grow_int + data_list$grow_z * z, data_list$grow_sd)
pb_z <- function(z) 1 / (1 + exp(-(data_list$repr_int + data_list$repr_z * z)))
pr_z <- function(Nt) 1 / (1 + exp(-(data_list$recr_int + data_list$recr_Nt * Nt)))
c_z1z <- function(z_prime, z, Nt) dnorm(z_prime, data_list$rcsz_int + data_list$rcsz_z * z + data_list$rcsz_Nt * Nt, data_list$rcsz_sd)
```

### Visualize DD Effects

```{r}
h <- (U - L) / m
meshpts <- L + ((1:m) - 1/2) * h
N_vals <- c(50, 200, 323, 450)
z_plot <- seq(1.5, 3.5, length.out = 100)
cols <- rainbow(length(N_vals))

par(mfrow = c(1, 2))

plot(NULL, xlim = range(z_plot), ylim = c(0, 1),
     xlab = "Size", ylab = "s(z)", main = "Survival vs size at different N")
for (i in seq_along(N_vals)) {
  lines(z_plot, sapply(z_plot, s_z, Nt = N_vals[i]), lwd = 2, col = cols[i])
}
legend("topleft", paste("N =", N_vals), col = cols, lwd = 2, cex = 0.8)

N_range <- seq(0, 500, length.out = 100)
plot(N_range, sapply(N_range, pr_z), type = "l", lwd = 2,
     xlab = "N", ylab = expression(p[r](N)),
     main = "Recruitment probability vs N")
par(mfrow = c(1, 1))
```

## Building the DD Kernel (Manual Iteration)

For the density-dependent model, the kernel must be rebuilt at each time step based on the current population. We first build a function that constructs the kernel matrix at a given population density.

```{r}
build_K <- function(Nt) {
  P <- matrix(0, m, m)
  F_mat <- matrix(0, m, m)
  for (j in 1:m) {
    for (i in 1:m) {
      P[i, j] <- s_z(meshpts[j], Nt) * g_z1z(meshpts[i], meshpts[j]) * h
      F_mat[i, j] <- 0.5 * s_z(meshpts[j], Nt) * pb_z(meshpts[j]) *
        pr_z(Nt) * c_z1z(meshpts[i], meshpts[j], Nt) * h
    }
  }
  return(P + F_mat)
}
```

## Lambda as a Function of Population Size

At the equilibrium population size $\bar{N}$, the asymptotic growth rate $\lambda(\bar{N}) = 1$. We can compute $\lambda(N)$ for a range of $N$ values.

```{r}
lambda_at_N <- function(N) {
  K <- build_K(N)
  Re(eigen(K)$values[1])
}

N_test <- seq(100, 500, length.out = 50)
lam_test <- sapply(N_test, lambda_at_N)

plot(N_test, lam_test, type = "l", lwd = 2,
     xlab = "Total population N", ylab = expression(lambda(N)),
     main = "Growth rate vs population size")
abline(h = 1, col = "red", lty = 2, lwd = 2)
legend("topright", c(expression(lambda(N)), expression(lambda == 1)),
       lty = c(1, 2), col = c("black", "red"), lwd = 2)
```

## Finding the Equilibrium

The equilibrium $\bar{N}$ is the population size where $\lambda(N) = 1$. We find it using `uniroot`.

```{r}
root_fn <- function(N) lambda_at_N(N) - 1.0
result <- uniroot(root_fn, interval = c(200, 400), tol = 1e-6)
N_eq <- result$root
lam_eq <- lambda_at_N(N_eq)
```

```{r}
#| echo: false
cat("Equilibrium population: N̄ =", round(N_eq, 1), "females\n")
cat("Lambda at equilibrium:", round(lam_eq, 6), "\n")
```

## Population Trajectories

We can iterate the DD-IPM from different starting population sizes to see convergence to the equilibrium.

```{r}
iterate_dd <- function(N0, n_steps) {
  n_t <- rep(N0 / (m * h), m)
  Nt_series <- sum(n_t) * h

  for (t in 1:n_steps) {
    Nt <- sum(n_t) * h
    K <- build_K(Nt)
    n_t <- K %*% n_t
    Nt_series <- c(Nt_series, sum(n_t) * h)
  }
  return(Nt_series)
}

N0_vals <- c(50, 150, 250, 350, 450)
n_steps <- 20

plot(NULL, xlim = c(0, n_steps), ylim = c(0, 500),
     xlab = "Time step", ylab = "Total population N",
     main = "DD-IPM trajectories from different starting sizes")
cols <- rainbow(length(N0_vals))
for (i in seq_along(N0_vals)) {
  Nt <- iterate_dd(N0_vals[i], n_steps)
  lines(0:n_steps, Nt, lwd = 2, col = cols[i])
}
abline(h = N_eq, lty = 2, lwd = 2)
legend("topright",
       c(paste("N₀ =", N0_vals), paste("Equilibrium ≈", round(N_eq))),
       col = c(cols, "black"), lty = c(rep(1, length(N0_vals)), 2),
       lwd = 2, cex = 0.7)
```

All trajectories converge to the same equilibrium, regardless of starting population size.

## Using ipmr with DD

We can also build this model using the `ipmr` pipeline with `di_dd = "dd"`. In density-dependent models, `sum(n_z_t)` provides the total population size at each time step.

```{r}
dd_ipm <- init_ipm(sim_gen = "simple", di_dd = "dd", det_stoch = "det") %>%
  define_kernel(
    name          = "P",
    formula       = s * g,
    family        = "CC",
    s             = 1 / (1 + exp(-(surv_int + surv_z * z_1 + surv_Nt * sum(n_z_t) * d_z))),
    g             = dnorm(z_2, grow_mu, grow_sd),
    grow_mu       = grow_int + grow_z * z_1,
    data_list     = data_list,
    states        = list(c("z")),
    uses_par_sets = FALSE,
    evict_cor     = FALSE
  ) %>%
  define_kernel(
    name          = "F",
    formula       = 0.5 * s * pb_pr * c_z1z,
    family        = "CC",
    s             = 1 / (1 + exp(-(surv_int + surv_z * z_1 + surv_Nt * sum(n_z_t) * d_z))),
    pb_pr         = (1 / (1 + exp(-(repr_int + repr_z * z_1)))) *
                    (1 / (1 + exp(-(recr_int + recr_Nt * sum(n_z_t) * d_z)))),
    c_z1z         = dnorm(z_2, rcsz_mu, rcsz_sd),
    rcsz_mu       = rcsz_int + rcsz_z * z_1 + rcsz_Nt * sum(n_z_t) * d_z,
    data_list     = data_list,
    states        = list(c("z")),
    uses_par_sets = FALSE,
    evict_cor     = FALSE
  ) %>%
  define_impl(
    make_impl_args_list(
      kernel_names = c("P", "F"),
      int_rule     = rep("midpoint", 2),
      state_start  = rep("z", 2),
      state_end    = rep("z", 2)
    )
  ) %>%
  define_domains(z = c(L, U, m)) %>%
  define_pop_state(n_z = rep(50 / (m * h), m)) %>%
  make_ipm(iterate = TRUE, iterations = 30, normalize_pop_size = FALSE)
```

```{r}
total_pop <- sapply(dd_ipm$pop_state$n_z, function(x) sum(x) * h)
all_lambdas <- lambda(dd_ipm, type_lambda = "all")

par(mfrow = c(1, 2))
plot(total_pop, type = "l", lwd = 2,
     xlab = "Time step", ylab = "Total population N",
     main = "DD-IPM via ipmr")
abline(h = N_eq, col = "red", lty = 2, lwd = 2)

plot(all_lambdas, type = "l", lwd = 2,
     xlab = "Time step", ylab = expression(lambda(t)),
     main = "Per-step growth rate (DD model)")
abline(h = 1, col = "red", lty = 2, lwd = 2)
par(mfrow = c(1, 1))
```

## Equilibrium Size Distribution

```{r}
n_eq <- rep(N_eq / (m * h), m)
for (i in 1:50) {
  Nt <- sum(n_eq) * h
  K_eq <- build_K(Nt)
  n_eq <- K_eq %*% n_eq
}

n_eq_norm <- n_eq / sum(n_eq)

plot(meshpts, n_eq_norm, type = "l", lwd = 2,
     xlab = "Log body mass", ylab = "Proportion",
     main = paste0("Equilibrium size distribution (N̄ ≈ ", round(N_eq), ")"))
polygon(c(meshpts, rev(meshpts)),
        c(n_eq_norm, rep(0, m)),
        col = adjustcolor("blue", alpha.f = 0.3), border = NA)
lines(meshpts, n_eq_norm, lwd = 2)
```

## Summary

In this vignette we:

1. Built a density-dependent IPM where survival, recruitment, and recruit size decrease with population size
2. Computed $\lambda(N)$ across a range of population sizes
3. Found the equilibrium population $\bar{N} \approx 323$ females where $\lambda = 1$
4. Demonstrated convergence from multiple starting conditions
5. Used both manual iteration and the `ipmr` DD interface with `sum(n_z_t) * d_z`
